
---
title: 'Data 606 Lab5b - Foundations for statistical inference - Confidence intervals'
Author: 'Sufian'
output:
  html_document:
    css: ./lab.css
    highlight: pygments
    theme: cerulean
  pdf_document: default
---


Rpub link:

http://rpubs.com/ssufian/533918

Github link:

https://github.com/ssufian/Data_606


## Sampling from Ames, Iowa

If you have access to data on an entire population, say the size of every 
house in Ames, Iowa, it's straight forward to answer questions like, "How big 
is the typical house in Ames?" and "How much variation is there in sizes of 
houses?". If you have access to only a sample of the population, as is often 
the case, the task becomes more complicated. What is your best guess for the 
typical size if you only know the sizes of several dozen houses? This sort of 
situation requires that you use your sample to make inference on what your 
population looks like.

## The data

In the previous lab, ``Sampling Distributions'', we looked at the population data
of houses from Ames, Iowa. Let's start by loading that data set.

```{r load-data, eval=TRUE}
load("more/ames.RData")
head(ames)
```

In this lab we'll start with a simple random sample of size 60 from the 
population. Specifically, this is a simple random sample of size 60. Note that 
the data set has information on many housing variables, but for the first 
portion of the lab we'll focus on the size of the house, represented by the 
variable `Gr.Liv.Area`.

```{r sample, eval=TRUE}
population <- ames$Gr.Liv.Area
samp <- sample(population, 60)
head(samp)
```

1.  Describe the distribution of your sample. What would you say is the 
"typical" size within your sample? Also state precisely what you interpreted 
"typical" to mean.
```{r}
hist(samp)
summary(samp)
sd(samp)
```

--------------------------------------------------------------------------------

\clearpage

ans:

The distribution is nearly normal with a right skew to it.  The Typical size is around 1250.  By eyballing

it, the number with the highest frequency is around 1250.  This could be interpreted as the average.  


--------------------------------------------------------------------------------

\clearpage



2.  Would you expect another student's distribution to be identical to yours? 
Would you expect it to be similar? Why or why not?

```{r } 
hist(samp,breaks=15, probability = TRUE)
meanarea = mean(samp)
sdarea = sd(samp)

x <- 450:3600
y <- dnorm(x = x, mean = meanarea , sd = sdarea)
lines(x = x, y = y, col = "red")

qqnorm(samp)
qqline(samp)

mean(population)
sd(population)
```

--------------------------------------------------------------------------------

\clearpage

ans:

Not identical but similar is expected.  This distribution borne out of sample size of 60 shows. It is

quite normal but right skewed.  The population mean is not too far off from the sample and Std. Dev. is

higher than the sample but not too terribly far off as well.

--------------------------------------------------------------------------------

\clearpage


## Confidence intervals

One of the most common ways to describe the typical or central value of a 
distribution is to use the mean. In this case we can calculate the mean of the 
sample using,

```{r sample-mean, eval=TRUE}
sample_mean <- mean(samp)

sample_mean
```

Return for a moment to the question that first motivated this lab: based on 
this sample, what can we infer about the population? Based only on this single 
sample, the best estimate of the average living area of houses sold in Ames 
would be the sample mean, usually denoted as $\bar{x}$ (here we're calling it 
`sample_mean`). That serves as a good *point estimate* but it would be useful 
to also communicate how uncertain we are of that estimate. This can be 
captured by using a *confidence interval*.

We can calculate a 95% confidence interval for a sample mean by adding and 
subtracting 1.96 standard errors to the point estimate (See Section 4.2.3 if 
you are unfamiliar with this formula).

```{r ci, eval=TRUE}
se <- sd(samp) / sqrt(60)
lower <- sample_mean - 1.96 * se
upper <- sample_mean + 1.96 * se
c(lower, upper)
```

This is an important inference that we've just made: even though we don't know 
what the full population looks like, we're 95% confident that the true 
average size of houses in Ames lies between the values *lower* and *upper*. 
There are a few conditions that must be met for this interval to be valid.

3.  For the confidence interval to be valid, the sample mean must be normally 
distributed and have standard error $s / \sqrt{n}$. What conditions must be 
met for this to be true?

--------------------------------------------------------------------------------

\clearpage

ans:

The central limit theorem is needed for confidence intervals to be valid.   However, it is also necessary

that the data be collected from random samples.  Confidence intervals will not remedy poorly collected

data. Lastly, the larger the sample size the better, but the common practice in real life is n >=30 should

be enough.

--------------------------------------------------------------------------------

\clearpage


## Confidence levels

4.  What does "95% confidence" mean? If you're not sure, see Section 4.2.2.

--------------------------------------------------------------------------------

\clearpage

ans:

95% confidence means that we used a procedure that works 95% of the time to get this interval.  

That is, 95% of all intervals produced by the procedure will contain their corresponding parameters.  

For any one particular interval, the true population percentage is either inside the interval or outside 

the interval.  

--------------------------------------------------------------------------------

\clearpage

In this case we have the luxury of knowing the true population mean since we 
have data on the entire population. This value can be calculated using the 
following command:

```{r pop-mean, eval=TRUE}
mean(population)
```

5.  Does your confidence interval capture the true average size of houses in 
Ames? If you are working on this lab in a classroom, does your neighbor's 
interval capture this value? 

--------------------------------------------------------------------------------

\clearpage

ans:

Yes it does, The confidence interval ranges from 1312 to 1576.001.  However our sample mean is 1444

which is within our interval.  Now the population mean is also 1444 which is also within this interval.  


--------------------------------------------------------------------------------

\clearpage


6.  Each student in your class should have gotten a slightly different 
confidence interval. What proportion of those intervals would you expect to 
capture the true population mean? Why? If you are working in this lab in a 
classroom, collect data on the intervals created by other students in the 
class and calculate the proportion of intervals that capture the true 
population mean.

\clearpage

ans:

sampling my confidence interval, I found it to be between 1325 to 1613; which still captures my sample

mean of 1444.  And I would expect that 95% of the time, my interval will capture the true population mean

as well


--------------------------------------------------------------------------------

\clearpage


```{r}
#Create empty vectors where we can save the means and standard deviations that will be calculated from each sample.
samp_mean <- rep(NA,50)
samp_sd <- rep(NA, 50)
n <- 50

for (i in 1:50){
  samp <- sample(population, n) #obtain a sample of size n = 50 from the population
  samp_mean[i] <- mean(samp) #save sample mean in ith element of samp_mean
  samp_sd[i] <- sd(samp) #save sample sd in ith element of samp_sd
}

#length(samp_mean)
```


```{r}
#Construct the confidence intervals
lower_1 <- samp_mean - 1.96 * samp_sd / sqrt(n)
upper_1<- samp_mean + 1.96 * samp_sd/ sqrt(n)
```

```{r}
#View the first interval
lower_1[i]
upper_1[i]
```

--------------------------------------------------------------------------------

\clearpage


Using R, we're going to recreate many samples to learn more about how sample 
means and confidence intervals vary from one sample to another. *Loops* come 
in handy here (If you are unfamiliar with loops, review the [Sampling Distribution Lab](http://htmlpreview.github.io/?https://github.com/andrewpbray/oiLabs/blob/master/sampling_distributions/sampling_distributions.html)).

Here is the rough outline:

-   Obtain a random sample.
-   Calculate and store the sample's mean and standard deviation.
-   Repeat steps (1) and (2) 50 times.
-   Use these stored statistics to calculate many confidence intervals.


But before we do all of this, we need to first create empty vectors where we 
can save the means and standard deviations that will be calculated from each 
sample. And while we're at it, let's also store the desired sample size as `n`.

```{r set-up, eval=TRUE}
samp_mean <- rep(NA, 50)
samp_sd <- rep(NA, 50)
n <- 60
```

Now we're ready for the loop where we calculate the means and standard deviations of 50 random samples.

```{r loop, eval=TRUE, tidy = FALSE}
for(i in 1:50){
  samp <- sample(population, n) # obtain a sample of size n = 60 from the population
  samp_mean[i] <- mean(samp)    # save sample mean in ith element of samp_mean
  samp_sd[i] <- sd(samp)        # save sample sd in ith element of samp_sd
}
```

Lastly, we construct the confidence intervals.

```{r ci50, eval=TRUE}
lower_vector <- samp_mean - 1.96 * samp_sd / sqrt(n) 
upper_vector <- samp_mean + 1.96 * samp_sd / sqrt(n)
```

Lower bounds of these 50 confidence intervals are stored in `lower_vector`, 
and the upper bounds are in `upper_vector`. Let's view the first interval.

```{r first-interval, eval=TRUE}
c(lower_vector[1], upper_vector[1])
```

* * *

## On your own

-   Using the following function (which was downloaded with the data set), 
    plot all intervals. What proportion of your confidence intervals include 
    the true population mean? Is this proportion exactly equal to the 
    confidence level? If not, explain why.

--------------------------------------------------------------------------------

\clearpage

ans:

In this particular sampling run, 3% of the proportion lies outside of the confidence intervals 

This is not exactly to the confidence level we would should expect but that maybe due to margin of error

sampling

--------------------------------------------------------------------------------

\clearpage


    ```{r plot-ci, eval=TRUE}
    plot_ci(lower_vector, upper_vector, mean(population))
    ```
```{r}
Vector <- data.frame(lower_vector, upper_vector)
meanp <- mean(population)

left <- sum(Vector$upper_vector < meanp)
right <- sum(Vector$lower_vector > meanp)

noMeanIncluded <- left + right
noMeanIncluded

proportion <- round(noMeanIncluded/n ,2)
proportion
```

-   Pick a confidence level of your choosing, provided it is not 95%. What is 
    the appropriate critical value?
    
--------------------------------------------------------------------------------

\clearpage


For a 90% confidence:

$z^⋆=1.65$

for a 99% confidence:

$z^⋆=1.58$


--------------------------------------------------------------------------------

\clearpage

    
```{r}
lower_vector2 <- samp_mean - 1.65 * samp_sd / sqrt(n) 
upper_vector2 <- samp_mean + 1.65 * samp_sd / sqrt(n)



lower_vector2[1] 
upper_vector[1] 
```
    

-   Calculate 50 confidence intervals at the confidence level you chose in the 
    previous question. You do not need to obtain new samples, simply calculate 
    new intervals based on the sample means and standard deviations you have 
    already collected. Using the `plot_ci` function, plot all intervals and 
    calculate the proportion of intervals that include the true population 
    mean. How does this percentage compare to the confidence level selected for
    the intervals?
```{r}
plot_ci(lower_vector2, upper_vector2, mean(population))
```

--------------------------------------------------------------------------------

\clearpage

ans:

Again, due to uncertainty in sampling, 98% include the population mean. This proportion is not necesarily

the same as our confidence level but actually better; yet again due to sampling error 


--------------------------------------------------------------------------------

\clearpage
